# Analysis of U.S. Weather Events (1996-2011)

## Synopsis
The purpose of this report is to analyze data from the [NOAA Storm Database](http://www.ncdc.noaa.gov/stormevents/) and determine which events are most harmful to population health and which events have the greatest economic consequences.

## Data Processing
The raw data used for this analysis was downloaded from the following url:
    [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
    
#### Reading in the data
I first read in the entire data set directly from the zipped csv file provided at the link above. The file contains 37 columns, eight of which we need to perform our analysis. Therefore we ignore all other columns and reset the column names to a more readable format. I read the date column in as a year value and ignore any other information. The event type is converted to all uppercase as it's read in.

<table cellspacing="10">
    <tr>
        <th>Original Column Name</th>
        <th>Formatted Column Name</th>
        <th>Description</th>
    </tr>
    <tr>
        <td>BGN_DATE</td>
        <td>year</td>
        <td>The year the weather event occurred.</td>
    </tr>
    <tr>
        <td>EVTYPE</td>
        <td>type</td>
        <td>The type of the weather event.</td>
    </tr>
    <tr>
        <td>FATALITIES</td>
        <td>fatalities</td>
        <td>The number of fatalities caused by the weather event.</td>
    </tr>
    <tr>
        <td>INJURIES</td>
        <td>injuries</td>
        <td>The number of injuries caused by the weather event.</td>
    </tr>
    <tr>
        <td>PROPDMG</td>
        <td>propertyDamage
        <td>The cost in dollars of property damage caused by the weather event.<td>
    </tr>
    <tr>
        <td>PROPDMGEXP</td>
        <td>propertyDamageScale</td>
        <td>The scale of the property damage caused by the weather event.
            <ul>
                <li>K = thousands of dollars</li>
                <li>M = millions of dollars</li>
                <li>B = billions of dollars</li>
            </ul>
        </td>
    </tr>
    <tr>
        <td>CROPDMG</td>
        <td>cropDamage
        <td>The cost in dollars of crop damage caused by the weather event.<td>
    </tr>
    <tr>
        <td>CROPDMGEXP</td>
        <td>cropDamageScale</td>
        <td>The scale of the crop damage caused by the weather event.
            <ul>
                <li>K = thousands of dollars</li>
                <li>M = millions of dollars</li>
                <li>B = billions of dollars</li>
            </ul>
        </td>
    </tr> 
</table>

```{r readInData, cache=TRUE}
# Set up date processing and convert all event types to upper case
library("lubridate")
setClass("myDate")
setClass("typeUpper")
setAs("character", "myDate", function(from) as.numeric(year(mdy_hms(from))))
setAs("character", "typeUpper", function(from) toupper(from))


# Filter out unnecessary columns
columnsToRead = c("NULL",
                  "myDate",
                  rep("NULL", 5),
                  "typeUpper",
                  rep("NULL", 14),
                  "numeric",    "numeric",  "numeric",   "character",  "numeric",    "character",
                  rep("NULL", 9))

# read data
eventData = read.csv("repdata-data-StormData.csv.bz2", header=TRUE, colClasses=columnsToRead, comment.char="")
colnames(eventData) <- 
 c("year", "type", "fatalities", "injuries", "propertyDamage", "propertyDamageScale", "cropDamage", "cropDamageScale")
```


#### Cleaning the data
There are several things I do to clean up the raw data set. First, I filter out all weather events that happened before 1996. According to the [NOAA](http://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype), 1996 is the year that information on all event types started to be collected. It also makes sense to limit the data to recent years as weather patterns change and dollar values in the data have not been adjusted for inflation.

``` {r filterEarlyYears, cache=TRUE, dependson="readInData"}
eventData <- eventData[eventData$year >= 1996, ]
```

Next I eliminate events that do not have any fatalities/injuries and property/crop damage.

``` {r removeEventsWithoutEffects, cache=TRUE, dependson="filterEarlyYears"}
eventData <- eventData[eventData$fatalities != 0 | eventData$fatalities != 0 | 
                       eventData$propertyDamage != 0 | eventData$cropDamage != 0, ]
```

Next I replace each pair of property and crop damage columns with a single column that has the numeric value representing each pair.
``` {r setNumericValues, cache=TRUE, dependson="removeEventsWithoutEffects"}
# return actual numeric value for damage data
# expects a two-column dataframe that contains the damage and damage scale columns
getNumericValue <- function(damageColumns)
{
    if ("" == damageColumns[2]) return(as.numeric(damageColumns[1]))
    if ("K" == damageColumns[2]) return(1000 * as.numeric(damageColumns[1]))
    if ("M" == damageColumns[2]) return(1000000 * as.numeric(damageColumns[1]))
    if ("B" == damageColumns[2]) return(1000000000 * as.numeric(damageColumns[1]))
}

propertyDamage <- apply(eventData[, c("propertyDamage", "propertyDamageScale")], 1, getNumericValue)
cropDamage <- apply(eventData[, c("cropDamage", "cropDamageScale")], 1, getNumericValue)
eventData <- cbind(eventData[, 1:4], propertyDamage, cropDamage)
```

The data now has the following format:
``` {r showDataFormat, dependson="setNumericValues"}
head(eventData)
```

#### Cleaning Event Types
The NOAA documentation specifies the following 48 event types:

``` {r validEventTypes, cache=TRUE}
validEventTypes <-
    as.vector(sapply(c("Astronomical Low Tide",   "Avalanche",           "Blizzard",
                       "Coastal Flood",           "Cold/Wind Chill",     "Debris Flow", 
                       "Dense Fog",               "Dense Smoke",         "Drought",
                       "Dust Devil",              "Dust Storm",          "Excessive Heat",
                       "Extreme Cold/Wind Chill", "Flash Flood",         "Flood",
                       "Frost/Freeze",            "Funnel Cloud",        "Freezing Fog",
                       "Hail",                    "Heat",                "Heavy Rain",
                       "Heavy Snow",              "High Surf",           "High Wind",
                       "Hurricane (Typhoon)",     "Ice Storm",           "Lake-Effect Snow",
                       "Lakeshore Flood",         "Lightning",           "Marine Hail",
                       "Marine High Wind",        "Marine Strong Wind",  "Marine Thunderstorm Wind",
                       "Rip Current",             "Seiche",              "Sleet",
                       "Storm Surge/Tide",        "Strong Wind",         "Thunderstorm Wind",
                       "Tornado",                 "Tropical Depression", "Tropical Storm",
                       "Tsunami",                 "Volcanic Ash",        "Waterspout",
                       "Wildfire",                "Winter Storm",        "Winter Weather"), toupper))
```


After displaying the percentage of rows that contain invalid event values and those invalid values, it is clear we need to clean up the event type data.
``` {r displayBadEventTypes, cache=TRUE, dependson="setNumericValues"}
invalidEventTypes <- eventData[!(eventData$type %in% validEventTypes), ]
print(paste("The percentage of events that have invalid event types = ", 
            round(nrow(invalidEventTypes) / nrow(eventData) * 100, 2)))
print(unique(invalidEventTypes$type))
```

Classifying invalid event types is difficult and cannot be done perfectly. For example, does "GUSTY WIND/HAIL" get classified under a wind event type or hail type? I reviewed the invalid event types and classified them as best I could with the rules listed below. After making the substitutions, I then removed any events classified as "OTHER" (both those originally classified as "OTHER" and ones I classified as "OTHER").

``` {r cleanEventTypes, cache=TRUE, dependson="displayBadEventTypes"}
convertToValidEventType <- function(eventType)
{
    if (grepl("MARINE TSTM WIND", eventType)) return("MARINE THUNDERSTORM WIND")
    if (grepl("TSTM WIND", eventType)) return("THUNDERSTORM WIND")
    if (grepl("THUNDERSTORM", eventType)) return("THUNDERSTORM WIND")
    if (grepl("GUSTY WIND", eventType)) return("HIGH WIND")
    if (!(grepl("^EXTREME", eventType)) & grepl("COLD", eventType)) return("COLD/WIND CHILL")
    if (grepl("^EXTREME", eventType)) return("EXTREME COLD/WIND CHILL")
    if (grepl("EXPOSURE", eventType)) return("EXTREME COLD/WIND CHILL")
    if (grepl("^FOG", eventType)) return("DENSE FOG")
    if (grepl("ROAD", eventType)) return("WINTER WEATHER")
    if (grepl("WEATHER", eventType)) return("WINTER WEATHER")
    if (grepl("BLACK ICE", eventType)) return("WINTER WEATHER")
    if (grepl("LAKE EFFECT SNOW", eventType)) return("LAKE-EFFECT SNOW")
    if (!(grepl("^LAKE", eventType)) & grepl("SNOW", eventType)) return("HEAVY SNOW")        
    if (grepl("SURF", eventType)) return("HIGH SURF")
    if (grepl("FREEZING SPRAY", eventType)) return("HIGH SURF")
    if (grepl("FREEZING RAIN", eventType)) return("SLEET")
    if (grepl("MIX", eventType)) return("SLEET")
    if (grepl("FREEZ", eventType) & !(grepl("FOG", eventType))) return("FROST/FREEZE")
    if (grepl("FROST", eventType)) return("FROST/FREEZE")
    if (grepl("ASTRONOMICAL", eventType)) return("ASTRONOMICAL LOW TIDE")
    if (grepl("COAST", eventType)) return("COASTAL FLOOD")
    if (grepl("EROSION", eventType)) return("COASTAL FLOOD")
    if (grepl("FLASH", eventType)) return("FLASH FLOOD")
    if (grepl("HURRICANE", eventType)) return("HURRICANE (TYPHOON)")
    if (grepl("TYPHOON", eventType)) return("HURRICANE (TYPHOON)")
    if (grepl("CURRENTS", eventType)) return("RIP CURRENT")
    if (grepl("FIRE", eventType)) return("WILDFIRE")
    if (grepl("TIDAL", eventType)) return("STORM SURGE/TIDE")
    if (grepl("SURGE", eventType)) return("STORM SURGE/TIDE")
    if (grepl("BURST", eventType)) return("HEAVY RAIN")
    if (grepl("RAIN", eventType)) return("HEAVY RAIN")
    if (grepl(" SEAS", eventType)) return("MARINE HIGH WIND")
    if (grepl("SWELLS", eventType)) return("MARINE HIGH WIND")
    if (grepl("WIND AND WAVE", eventType)) return("MARINE HIGH WIND")
    if (grepl("RIVER", eventType)) return("FLOOD")
    if (grepl("DAM", eventType)) return("FLOOD")
    if (grepl("HIGH WATER", eventType)) return("FLOOD")
    if (grepl("HEAT", eventType)) return("EXCESSIVE HEAT")
    if (grepl("BLOWING DUST", eventType)) return("DUST STORM")
    if (grepl("LANDSPOUT", eventType)) return("DUST DEVIL")
    if (grepl("^WIND", eventType)) return("HIGH WIND")
    if (grepl("WINDS", eventType)) return("HIGH WIND")
    if (grepl("GRADIENT WIND", eventType)) return("HIGH WIND")
    if (grepl("G40", eventType)) return("HIGH WIND")
    if (grepl("WHIRLWIND", eventType)) return("TORNADO")
    if (grepl("SMALL HAIL", eventType)) return("HAIL")
    if (grepl("ICE JAM", eventType)) return("OTHER")
    if (grepl("URBAN", eventType)) return("OTHER")
    if (grepl("SLIDE", eventType)) return("OTHER")
    if (grepl("SLUMP", eventType)) return("OTHER")
    if (grepl("GLAZE", eventType)) return("OTHER")
    if (grepl("DROWN", eventType)) return("OTHER")
    if (grepl("UNSEASONABLY", eventType)) return("OTHER")
    if (grepl("ACCIDENT", eventType)) return("OTHER")
    
    return(eventType)
}

eventData$type <- sapply(eventData$type, convertToValidEventType)
eventData <- eventData[eventData$type != "OTHER", ]
```

#### Outliers
Finally, I checked to see if there were any outliers in the data. I generated box plots for all of the values:

``` {r findOutliers, cache=TRUE, dependson="cleanEventTypes"}
par(mfrow = c(1, 2))
boxplot(eventData$fatalities, eventData$injuries, 
        main="Distribution of Fatalities/Injuries",
        xlab="Fatalities | Injuries")

boxplot(eventData$propertyDamage, eventData$cropDamage,
        main="Distribution of Property/Crop Damage",
        xlab="Property | Crop",
        ylab="Number of Dollars")
```

It appears there is an outlier in the property damage for a singe event. It appears likely that the data set records the damame in billions rather than millions. I determined the event for this oulier and converted the property damage to millions of dollars.

``` {r fixOutlier, cache=TRUE, dependson="findOutliers"}
maxIndex = which.max(eventData$propertyDamage)
print(eventData[maxIndex,])
eventData$propertyDamage[maxIndex] <- eventData$propertyDamage[maxIndex] / 1000
```

## Analysis

First draw a graph of the fatalities

``` {r humanHealthPlot, dependson="fixOutlier"}
library(ggplot2)
eventData$type <- factor(eventData$type)
healthData <- setNames(aggregate(eventData$fatalities + eventData$injuries, 
                                 by=eventData[c("type")], 
                                 FUN=sum), 
                       c("type", "amount"))
healthData <- healthData[order(-healthData$amount),][1:10,]

ggplot(healthData,
       aes(reorder(healthData$type, order(healthData$amount, decreasing=TRUE)), 
           healthData$amount)) +
       ggtitle("Effects of Weather Events on Population Health") +
       xlab("Event Type") + 
       ylab("Number of Fatalities and Injuries") +
       geom_bar(stat="identity") +  
       theme(axis.text.x = element_text(angle = -45, hjust = 0)) 
```

Now draw a graph of the damage

``` {r damagePlot, dependson="fixOutlier"}
damageData <- setNames(aggregate(eventData$propertyDamage + eventData$cropDamage, 
                                 by=eventData[c("type")], 
                                 FUN=sum), 
                       c("type", "amount"))
damageData <- damageData[order(-damageData$amount),][1:10,]

ggplot(damageData,
       aes(reorder(damageData$type, order(damageData$amount, decreasing=TRUE)), 
           damageData$amount)) +
       ggtitle("Effects of Weather Events on Population Health") +
       xlab("Event Type") + 
       ylab("Number of Fatalities and Injuries") +
       geom_bar(stat="identity") +  
       theme(axis.text.x = element_text(angle = -45, hjust = 0))
```